---
title: "Comparison with PySTAC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with PySTAC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(stacbuildr)
library(reticulate)

py_require("pystac")
pystac <- import("pystac")
dt <- import("datetime")
```

# Comparison with PySTAC

This vignette compares stacbuildr with the Python
[PySTAC](https://pystac.readthedocs.io/) library by writing equivalent catalog
structures to disk and comparing the resulting JSON.

## Building a catalog with an item

### stacbuildr

```{r}
catalog <- stac_catalog(
  id = "environmental-data",
  description = "A collection of environmental data",
  title = "Environmental Data Catalog"
)

item <- stac_item(
  id = "temperature",
  geometry = list(
    type = "Point",
    coordinates = c(-122.45, 54.34)
  ),
  bbox = c(-125, 50, -120, 55),
  datetime = "2023-07-06T08:00:00Z"
)

catalog <- catalog |> 
  add_item(item)
```

### PySTAC

```{r}
catalog_py <- pystac$Catalog(
  id = "environmental-data",
  description = "A collection of environmental data",
  title = "Environmental Data Catalog",
  catalog_type = pystac$CatalogType$RELATIVE_PUBLISHED
)

item_py <- pystac$Item(
  id = "temperature",
  geometry = reticulate::dict(
    type = "Point",
    coordinates = list(c(-122.45, 54.34))
  ),
  bbox = list(-125, 50, -120, 55),
  datetime = dt$datetime$fromisoformat("2023-07-06T08:00:00+00:00"),
  properties = reticulate::dict()
)

catalog_py$add_item(item_py)
```

## Writing to disk and comparing

Both libraries are asked to write a relative catalog. stacbuildr uses
`catalog_type = "relative"` and PySTAC uses `CatalogType$RELATIVE_PUBLISHED`,
which are equivalent.

```{r}
r_path <- file.path(tempdir(), "r-stac")
write_stac(catalog, r_path, catalog_type = "relative", overwrite = TRUE)
catalog_rjson <- jsonlite::read_json(file.path(r_path, "catalog.json"))

py_path <- file.path(tempdir(), "py-stac")
catalog_py$normalize_hrefs(py_path)
catalog_py$save(dest_href = py_path)
catalog_pyjson <- jsonlite::read_json(file.path(py_path, "catalog.json"))
```

### Catalog JSON

```{r}
cat("R json output: \n")
jsonlite::toJSON(catalog_rjson, pretty = TRUE, auto_unbox = TRUE)

cat("\nPython json output: \n")
jsonlite::toJSON(catalog_pyjson, pretty = TRUE, auto_unbox = TRUE)
```

### Item JSON

```{r}
r_item <- jsonlite::read_json(file.path(r_path, "temperature", "temperature.json"))
py_item <- jsonlite::read_json(file.path(py_path, "temperature", "temperature.json"))

jsonlite::toJSON(r_item, pretty = TRUE, auto_unbox = TRUE)
jsonlite::toJSON(py_item, pretty = TRUE, auto_unbox = TRUE)
```

## Assets

The item assets are functionally equivalent, with the same fields and values.

```{r}
catalog <- stac_catalog(
  id = "environmental-data",
  description = "A collection of environmental data",
  title = "Environmental Data Catalog"
)

item <- stac_item(
  id = "temperature",
  geometry = list(
    type = "Point",
    coordinates = c(-122.45, 54.34)
  ),
  bbox = c(-125, 50, -120, 55),
  datetime = "2023-07-06T08:00:00Z"
)

asset <- stac_asset(
  href = "https://example.com/temperature.tif",
  title = "Temperature GeoTIFF",
  description = "A GeoTIFF file containing temperature data.",
  type = "image/tiff; application=geotiff"
)

item <- item |> 
  add_asset("temperature", asset)

catalog <- catalog |> 
  add_item(item)
```


```{r}
asset_py <- pystac$Asset(
  href = "https://example.com/temperature.tif",
  title = "Temperature GeoTIFF",
  description = "A GeoTIFF file containing temperature data.",
  media_type = "image/tiff; application=geotiff"
)

item_py$add_asset("temperature", asset_py)
```

Write the catalogs to disk again to include the asset and read the item JSON:

```{r}
write_stac(catalog, r_path, catalog_type = "relative", overwrite = TRUE)
catalog_rjson <- jsonlite::read_json(file.path(r_path, "catalog.json"))

catalog_py$normalize_hrefs(py_path)
catalog_py$save(dest_href = py_path)
catalog_pyjson <- jsonlite::read_json(file.path(py_path, "catalog.json"))
```

Compare the asset in the resulting item JSON:

```{r}
r_item <- jsonlite::read_json(file.path(r_path, "temperature", "temperature.json"))
py_item <- jsonlite::read_json(file.path(py_path, "temperature", "temperature.json"))

cat("R json output: \n")
jsonlite::toJSON(r_item, pretty = TRUE, auto_unbox = TRUE)

cat("\nPython json output: \n")
jsonlite::toJSON(py_item, pretty = TRUE, auto_unbox = TRUE)
```