---
title: "Comparison with PySTAC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with PySTAC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(stacbuildr)
library(reticulate)

py_require("pystac")
pystac <- import("pystac")
```

# Comparison with PySTAC

This vignette compares stacbuildr with the Python
[PySTAC](https://pystac.readthedocs.io/) library by writing equivalent catalog
structures to disk and comparing the resulting JSON.

## Building a catalog with an item

### stacbuildr

```{r}
catalog <- stac_catalog(
  id = "environmental-data",
  description = "A collection of environmental data",
  title = "Environmental Data Catalog"
)

item <- stac_item(
  id = "temperature",
  geometry = list(
    type = "Point",
    coordinates = c(-122.45, 54.34)
  ),
  bbox = c(-125, 50, -120, 55),
  datetime = "2023-07-06T08:00:00Z"
)

catalog_r <- catalog |> add_item(item)
```

### PySTAC

```{r}
catalog_py <- pystac$Catalog(
  id = "environmental-data",
  description = "A collection of environmental data",
  title = "Environmental Data Catalog",
  catalog_type = pystac$CatalogType$RELATIVE_PUBLISHED
)

dt <- import("datetime")
item_py <- pystac$Item(
  id = "temperature",
  geometry = reticulate::dict(
    type = "Point",
    coordinates = list(c(-122.45, 54.34))
  ),
  bbox = list(-125, 50, -120, 55),
  datetime = dt$datetime$fromisoformat("2023-07-06T08:00:00+00:00"),
  properties = reticulate::dict()
)

catalog_py$add_item(item_py)
```

## Writing to disk and comparing

Both libraries are asked to write a relative catalog. stacbuildr uses
`catalog_type = "relative"` and PySTAC uses `CatalogType$RELATIVE_PUBLISHED`,
which are equivalent.

```{r}
r_path <- file.path(tempdir(), "r-stac")
write_stac(catalog_r, r_path, catalog_type = "relative", overwrite = TRUE)
r_catalog <- jsonlite::read_json(file.path(r_path, "catalog.json"))

py_path <- file.path(tempdir(), "py-stac")
catalog_py$normalize_hrefs(py_path)
catalog_py$save(dest_href = py_path)
py_catalog <- jsonlite::read_json(file.path(py_path, "catalog.json"))
```

### Catalog JSON

```{r}
cat("R json output: \n")
jsonlite::toJSON(r_catalog, pretty = TRUE, auto_unbox = TRUE)

cat("\nPython json output: \n")
jsonlite::toJSON(py_catalog, pretty = TRUE, auto_unbox = TRUE)
```

### Item JSON

```{r}
r_item <- jsonlite::read_json(file.path(r_path, "temperature", "temperature.json"))
py_item <- jsonlite::read_json(file.path(py_path, "temperature", "temperature.json"))

jsonlite::toJSON(r_item, pretty = TRUE, auto_unbox = TRUE)
jsonlite::toJSON(py_item, pretty = TRUE, auto_unbox = TRUE)
```

## Checks

The outputs are functionally equivalent with one known difference: pystac
produces an absolute path for the `self` link even in `RELATIVE_PUBLISHED`
mode, whereas stacbuildr uses a relative `./catalog.json`. This is a
deliberate behavioural difference in pystac.

```{r}
get_link <- function(links, rel) {
  Filter(function(x) x$rel == rel, links)[[1]]
}

# Top-level catalog fields match
stopifnot(all(names(py_catalog) %in% names(r_catalog)))

# Same number of links
stopifnot(length(py_catalog$links) == length(r_catalog$links))

# root link hrefs match
stopifnot(get_link(r_catalog$links, "root")$href == get_link(py_catalog$links, "root")$href)

# item link hrefs match
stopifnot(get_link(r_catalog$links, "item")$href == get_link(py_catalog$links, "item")$href)

# self link: stacbuildr uses relative, pystac uses absolute (known difference)
stopifnot(get_link(r_catalog$links, "self")$href == "./catalog.json")
stopifnot(startsWith(get_link(py_catalog$links, "self")$href, "/"))
```
