---
title: "Describing Data with STAC Extensions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Describing Data with STAC Extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The core STAC Item spec is deliberately small: an id, a geometry, a datetime,
and a bag of assets. Everything that makes a dataset *interpretable* — which
band is red, what the nodata value is, what pixel value 3 means, how to draw it
on a map — lives in **extensions**.

An extension is two things:

1. A schema URI added to the Item's `stac_extensions` array, which tells clients
   and validators which vocabulary is in play.
2. A set of prefixed fields (`eo:bands`, `raster:bands`, `classification:classes`,
   …) placed either in the Item's `properties` or on an individual asset.

`stacbuildr` gives each supported extension an `add_*_extension()` function that
handles both halves. You never edit `stac_extensions` by hand.

This vignette covers the extensions the package supports, when to reach for
each, and the recurring decision of item-level versus asset-level placement.

## Setup

```{r}
library(stacbuildr)
```

We will build up a single Landsat-like scene, adding one extension at a time.

```{r}
item <- stac_item(
  id = "LC09_L2SP_044033_20230615",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-122.5, 39.5), c(-120.5, 39.5), c(-120.5, 40.5),
      c(-122.5, 40.5), c(-122.5, 39.5)
    ))
  ),
  bbox = c(-122.5, 39.5, -120.5, 40.5),
  datetime = "2023-06-15T18:45:37Z",
  properties = list(
    platform = "landsat-9",
    instruments = c("oli", "tirs")
  )
)

item <- item |>
  add_asset(
    key = "red",
    href = "https://example.com/LC09_B4.tif",
    type = "image/tiff; application=geotiff; profile=cloud-optimized",
    roles = c("data")
  ) |>
  add_asset(
    key = "nir",
    href = "https://example.com/LC09_B5.tif",
    type = "image/tiff; application=geotiff; profile=cloud-optimized",
    roles = c("data")
  )
```

## Item-level or asset-level?

Nearly every `add_*_extension()` function takes an `asset_key` argument, and the
choice it controls matters more than it first appears.

* **Omit `asset_key`** and the fields land in `item@properties`. This says the
  metadata describes the whole scene.
* **Pass `asset_key`** and the fields land on that asset. This says the metadata
  describes only that file.

The rule of thumb: if the value would differ between two assets in the same
item, it belongs on the asset. Band descriptions, data types, nodata values and
resolutions are per-file, so they go on assets. Scene-wide facts — cloud cover,
the DOI of the paper describing the dataset — go on the item.

Passing an `asset_key` that does not exist is an error, so add assets before
extensions:

```{r, error = TRUE}
add_eo_extension(
  item,
  bands = list(eo_band(name = "B6", common_name = "swir16")),
  asset_key = "swir"
)
```

One field resists this choice: `table:storage_options` describes how to *open a
particular file*, so it only ever makes sense on an asset. `add_table_extension()`
requires an `asset_key` when you supply it.

## Electro-Optical (EO)

The [EO extension](https://github.com/stac-extensions/eo) describes the spectral
properties of imagery: what wavelength each band samples, and how much of the
scene was obscured by cloud or snow.

`eo_band()` builds one band; `add_eo_extension()` attaches a list of them along
with the optional scene-level `cloud_cover` and `snow_cover` percentages (both
validated to 0–100).

```{r}
item <- item |>
  add_eo_extension(cloud_cover = 12.5, snow_cover = 0) |>
  add_eo_extension(
    bands = list(
      eo_band(
        name = "B4",
        common_name = "red",
        center_wavelength = 0.655,
        full_width_half_max = 0.037
      )
    ),
    asset_key = "red"
  ) |>
  add_eo_extension(
    bands = list(
      eo_band(
        name = "B5",
        common_name = "nir",
        center_wavelength = 0.865,
        full_width_half_max = 0.028
      )
    ),
    asset_key = "nir"
  )

item@properties$`eo:cloud_cover`
item@assets$red$`eo:bands`
```

Note the pattern above: calling `add_eo_extension()` several times is fine. The
schema URI is only added once, and each call writes to a different place.

`common_name` is checked against the STAC common band names (`"coastal"`,
`"blue"`, `"green"`, `"red"`, `"rededge"`, `"nir"`, `"swir16"`, `"lwir11"`,
and so on). It is what lets a client ask for "the red band" without knowing that
Landsat calls it B4 and Sentinel-2 calls it B04.

### Sensor presets

Typing out band definitions for a well-known sensor is tedious and easy to get
wrong, so the package ships presets:

```{r}
bands <- landsat_oli_bands()
length(bands)
bands[[4]][c("name", "common_name", "center_wavelength")]
```

The others are `sentinel2_msi_bands()`, `worldview3_bands()`, `skysat_bands()`
and `planetscope_bands()`. `landsat_oli_bands()` takes
`include_thermal = TRUE` to append the TIRS bands.

## Raster

Where EO describes the *physics* of a band, the
[Raster extension](https://github.com/stac-extensions/raster) describes the
*pixels*: data type, nodata value, ground sample distance, and the scale/offset
needed to turn stored integers into physical units.

`raster_band()` is an S7 object rather than a plain list, so typos in
`data_type` are caught as they are written.

```{r}
red_band <- raster_band(
  nodata = 0,
  data_type = "uint16",
  spatial_resolution = 30,
  scale = 0.0000275,
  offset = -0.2,
  unit = "reflectance",
  statistics = raster_statistics(
    minimum = 1,
    maximum = 65455,
    mean = 9821,
    stddev = 3110,
    valid_percent = 98.7
  )
)

item <- add_raster_extension(item, bands = list(red_band), asset_key = "red")

item@assets$red$`raster:bands`[[1]]$scale
```

`raster_statistics()` and `raster_histogram()` build the optional `statistics`
and `histogram` sub-objects. Statistics are worth including: they let a client
pick sensible display limits without reading the file.

### Deriving bands from a file

Rather than transcribing metadata that GDAL already knows,
`raster_from_file()` reads it straight off the raster. With
`calculate_statistics = TRUE` it also samples the file to populate the
statistics block.

```{r}
tif <- system.file("ex/elev.tif", package = "terra")

if (requireNamespace("terra", quietly = TRUE) && nzchar(tif)) {
  bands <- raster_from_file(tif, calculate_statistics = TRUE)
  str(bands[[1]], max.level = 1)
}
```

`raster_from_file()` already returns a *list* of bands, one per layer. Pass it
straight through — wrapping it again in `list()` is the most common mistake here,
and the function raises a specific error when it detects the double-wrapping.

```{r, error = TRUE}
add_raster_extension(item, bands = list(raster_from_file(tif)))
```

`bands_from_terra()` does the same job for a `SpatRaster` you already have in
memory.

### EO and Raster together

The two extensions describe the same bands from different angles, and a single
band object can carry both vocabularies. Common fields (`name`, `description`,
`nodata`, `data_type`) are unprefixed; extension-specific ones are prefixed. Build
the band with `eo_band()` and pass the raster fields through `...`:

```{r}
combined <- eo_band(
  name = "B4",
  common_name = "red",
  center_wavelength = 0.655,
  nodata = 0,
  data_type = "uint16",
  "raster:spatial_resolution" = 30,
  "raster:scale" = 0.0000275
)

names(combined)
```

## Classification

The [Classification extension](https://github.com/stac-extensions/classification)
gives meaning to the integer values in a thematic raster. It has two mutually
exclusive modes, and `add_classification_extension()` errors if you supply both.

**Classes** map a pixel value to a name — land cover maps, crop type layers:

```{r}
lc_item <- stac_item(
  id = "nlcd-2021",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-122.5, 39.5), c(-120.5, 39.5), c(-120.5, 40.5),
      c(-122.5, 40.5), c(-122.5, 39.5)
    ))
  ),
  bbox = c(-122.5, 39.5, -120.5, 40.5),
  datetime = "2021-01-01T00:00:00Z"
) |>
  add_asset(
    key = "landcover",
    href = "https://example.com/nlcd-2021.tif",
    type = "image/tiff; application=geotiff",
    roles = c("data")
  )

classes <- list(
  classification_class(value = 0, name = "nodata", nodata = TRUE),
  classification_class(value = 11, name = "open-water",
                       title = "Open Water", color_hint = "466B9F"),
  classification_class(value = 42, name = "evergreen-forest",
                       title = "Evergreen Forest", color_hint = "1C5F2C"),
  classification_class(value = 82, name = "cultivated-crops",
                       title = "Cultivated Crops", color_hint = "AB6C28")
)

lc_item <- add_classification_extension(
  lc_item,
  classes = classes,
  asset_key = "landcover"
)
```

Two constraints catch mistakes early: `name` must be a slug (letters, digits,
hyphens, underscores — no spaces, which is why the titles above carry the
human-readable form), and `color_hint` must be exactly six upper-case hex digits
with no leading `#`.

```{r, error = TRUE}
classification_class(value = 11, name = "open water")
```

**Bitfields** describe QA bands where several flags are packed into one integer.
Each `classification_bitfield()` names a bit range by `offset` and `length` and
carries its own classes:

```{r}
qa_item <- add_classification_extension(
  lc_item,
  bitfields = list(
    classification_bitfield(
      offset = 0, length = 1, name = "fill",
      description = "Image or fill data",
      classes = list(
        classification_class(value = 0, name = "not-fill"),
        classification_class(value = 1, name = "fill")
      )
    ),
    classification_bitfield(
      offset = 3, length = 1, name = "cloud",
      description = "Cloud detected",
      classes = list(
        classification_class(value = 0, name = "not-cloud"),
        classification_class(value = 1, name = "cloud")
      )
    )
  )
)

vapply(qa_item@properties$`classification:bitfields`, `[[`, character(1), "name")
```

## Render

The [Render extension](https://github.com/stac-extensions/render) records how a
dataset is *meant* to be displayed — which assets to combine, how to stretch
them, which colormap to use. Dynamic tilers such as titiler read these
directly, so a viewer can produce the intended visualization without anyone
hardcoding parameters.

`renders` is a **named** list; the names become the keys under `renders` and are
what a client shows in a layer picker.

```{r}
item <- add_render_extension(
  item,
  renders = list(
    natural_color = render_object(
      assets = c("red", "nir"),
      title = "Natural Color",
      rescale = list(c(0, 10000), c(0, 10000)),
      resampling = "bilinear"
    ),
    ndvi = render_object(
      assets = c("nir", "red"),
      title = "NDVI",
      expression = "(nir-red)/(nir+red)",
      rescale = list(c(-1, 1)),
      colormap_name = "ylgn",
      minmax_zoom = c(6, 14)
    )
  )
)

names(item@properties$renders)
```

`rescale` is a list of two-element numeric vectors, one per band being
stretched — a flat `c(0, 10000)` is rejected. Render is the one extension here
that also accepts a `stac_collection()`, which is usually where it belongs: the
display recipe rarely varies scene to scene.

```{r}
collection <- stac_collection(
  id = "landsat-c2-l2",
  description = "Landsat Collection 2 Level-2 surface reflectance",
  license = "CC0-1.0",
  extent = stac_extent(
    spatial_bbox = list(c(-180, -90, 180, 90)),
    temporal_interval = list(list("2013-02-11T00:00:00Z", NULL))
  )
) |>
  add_render_extension(renders = list(
    natural_color = render_object(
      assets = c("red", "nir"),
      title = "Natural Color",
      rescale = list(c(0, 10000), c(0, 10000))
    )
  ))
```

## Scientific citation

The [Scientific extension](https://github.com/stac-extensions/scientific) records
how a dataset should be cited, and which publications describe it. Fields go in
the item properties — there is no `asset_key`, since a citation applies to the
dataset as a whole.

```{r}
item <- add_scientific_extension(
  item,
  doi = "10.5066/P9OGBGM6",
  citation = "U.S. Geological Survey (2023). Landsat 9 Collection 2 Level-2.",
  publications = list(
    scientific_publication(
      doi = "10.1016/j.rse.2020.111968",
      citation = "Masek, J. et al. (2020). Landsat 9: Empowering open science."
    )
  )
)

item@properties$`sci:doi`
```

`doi` must be a bare DOI name, not a URL — clients build the resolver link
themselves:

```{r, error = TRUE}
add_scientific_extension(item, doi = "https://doi.org/10.5066/P9OGBGM6")
```

## Table

The [Table extension](https://github.com/stac-extensions/table) describes tabular
assets — GeoParquet, CSV, Arrow — so a client knows the schema before
downloading. Columns are built with `table_column()`; `primary_geometry` names
the geometry column, and `storage_options` carries the arguments a reader needs
to open the file.

`asset_key` behaves as it does elsewhere: omit it and the columns describe the
item, supply it and they describe that one asset. The default — omitting it — places
 `table:columns`, `table:primary_geometry` and `table:row_count` under "Item Properties and
Collection Fields". That is the right choice for the common case of an item
wrapping a single table, so the example below omits it for those fields.

`storage_options` is the exception, and the one field `asset_key` is mandatory
for. The README gives it its own "Asset Object fields" section: it holds
fsspec-style arguments for opening one specific file, so it has no meaning at
item level.

```{r}
parcels <- stac_item(
  id = "parcels-2023",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-122.5, 39.5), c(-120.5, 39.5), c(-120.5, 40.5),
      c(-122.5, 40.5), c(-122.5, 39.5)
    ))
  ),
  bbox = c(-122.5, 39.5, -120.5, 40.5),
  datetime = "2023-01-01T00:00:00Z"
) |>
  add_asset(
    key = "data",
    href = "https://example.com/parcels.parquet",
    type = "application/x-parquet",
    roles = c("data")
  ) |>
  add_table_extension(
    columns = list(
      table_column(name = "geometry", type = "binary",
                   description = "Parcel boundary (WKB)"),
      table_column(name = "parcel_id", type = "int64"),
      table_column(name = "area_ha", type = "double"),
      table_column(name = "zoning", type = "string")
    ),
    primary_geometry = "geometry",
    row_count = 148302
  ) |>
  add_table_extension(
    storage_options = list(account_name = "parcelsdata"),
    asset_key = "data"
  )

vapply(parcels@properties$`table:columns`, `[[`, character(1), "name")
names(parcels@assets$data$`table:storage_options`)
```

When one item bundles several tables with different schemas, pass `asset_key`
and each asset carries its own column list:

```{r}
multi <- stac_item(
  id = "census-2021",
  geometry = list(type = "Point", coordinates = c(-121.5, 40)),
  bbox = c(-121.5, 40, -121.5, 40),
  datetime = "2021-01-01T00:00:00Z"
) |>
  add_asset(key = "households", href = "https://example.com/hh.parquet",
            type = "application/x-parquet") |>
  add_asset(key = "persons", href = "https://example.com/persons.parquet",
            type = "application/x-parquet") |>
  add_table_extension(
    columns = list(table_column(name = "household_id", type = "int64"),
                   table_column(name = "tenure", type = "string")),
    row_count = 24500,
    asset_key = "households"
  ) |>
  add_table_extension(
    columns = list(table_column(name = "person_id", type = "int64"),
                   table_column(name = "age", type = "int32")),
    row_count = 61200,
    asset_key = "persons"
  )

vapply(multi@assets$persons$`table:columns`, `[[`, character(1), "name")
```

One caveat on how much the spec backs any of this. The item/asset split is
stated only in the README's field tables; the extension's JSON schema reuses a
single permissive `fields` definition for properties, assets and `item_assets`
alike, so a validator will not object wherever you put the columns — and it does
not mention `table:storage_options` at all. Table is a *Pilot*-maturity
extension. Treat the placement above as the documented convention rather than
something a schema check enforces.

## Vector

The [Vector extension](https://github.com/stac-extensions/vector) records the
cartographic properties of a vector dataset: which geometry types it contains,
the minimum mapping unit and width used when it was generalized, and the scale
it was digitized for. These tell a user whether the data is fit for their
purpose — a layer captured at 1:50,000 should not be used for parcel-level work.

```{r}
parcels <- add_vector_extension(
  parcels,
  geometry_types = c("Polygon", "MultiPolygon"),
  mmu = 100,
  mmw = 5,
  reference_scale = 50000
)

parcels@properties[c("vector:geometry_types", "vector:reference_scale")]
```

`geometry_types` is checked against the seven GeoJSON geometry type names, so
lower-case or pluralized spellings fail rather than reaching a validator later.

Table and Vector are complementary and frequently appear together: Table
describes the columns, Vector describes the geometry.

## Datacube

The [Datacube extension](https://github.com/stac-extensions/datacube) describes
multidimensional arrays — NetCDF, Zarr, HDF — where the interesting structure is
the axes rather than the bands. It has two halves: `dimensions` names the axes,
and `variables` says which arrays exist over them.

Both are **named** lists, and the names in a variable's `dimensions` argument
must refer to dimensions you defined.

```{r}
cube <- stac_item(
  id = "era5-2023-06",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-180, -90), c(180, -90), c(180, 90), c(-180, 90), c(-180, -90)
    ))
  ),
  bbox = c(-180, -90, 180, 90),
  datetime = "2023-06-01T00:00:00Z"
) |>
  add_datacube_extension(
    dimensions = list(
      lon = cube_dimension(
        type = "spatial", axis = "x",
        extent = c(-180, 180), step = 0.25,
        reference_system = 4326
      ),
      lat = cube_dimension(
        type = "spatial", axis = "y",
        extent = c(-90, 90), step = 0.25,
        reference_system = 4326
      ),
      time = cube_dimension(
        type = "temporal",
        extent = c("2023-06-01T00:00:00Z", "2023-06-30T23:00:00Z"),
        step = "PT1H"
      ),
      level = cube_dimension(
        type = "vertical",
        values = c(1000, 850, 500, 250),
        unit = "hPa"
      )
    ),
    variables = list(
      temperature = cube_variable(
        type = "data",
        dimensions = c("lon", "lat", "time", "level"),
        unit = "K",
        data_type = "float32",
        description = "Air temperature"
      ),
      land_mask = cube_variable(
        type = "auxiliary",
        dimensions = c("lon", "lat"),
        data_type = "uint8"
      )
    )
  )

names(cube@properties$`cube:dimensions`)
```

A few points that are easy to get wrong:

* `type = "spatial"` requires an `axis` of `"x"`, `"y"` or `"z"`, and horizontal
  axes require a two-element `extent`.
* Use `extent` for a continuous range and `values` for an enumerated set — the
  pressure levels above are enumerated because they are not evenly spaced.
* `step` is a number for numeric axes and an ISO 8601 duration for temporal ones
  (`"PT1H"` is one hour). Set it to `NULL` for irregular spacing.
* Variables are `"data"` (the payload) or `"auxiliary"` (supporting arrays such
  as masks or coordinate lookups).

## Projection

Projection is the one extension with no `add_*` function, because it is applied
for you. `item_from_terra()` and `items_from_directory()` add the schema URI and
populate `proj:epsg`, `proj:wkt2`, `proj:shape`, `proj:bbox` and
`proj:transform` from the raster whenever its CRS is not WGS84. An item already
in EPSG:4326 needs none of it — the geometry and bbox say everything.

## Checking the result

Two levels of checking are available. `validate_stac()` runs the package's own
structural rules and needs no extra dependencies:

```{r}
result <- validate_stac(item)
result$valid
```

`validate_stac_schema()` goes further, fetching the official JSON schemas and
validating the serialized item against the core spec *and* against every schema
listed in `stac_extensions`. It requires the `jsonvalidate` package and network
access, which is why it is not run here:

```{r, eval = FALSE}
validate_stac_schema(item, validate_extensions = TRUE)
```

The second form is worth running once over a representative item before
publishing a catalog. Temper your expectations of it, though: extension schemas
vary a great deal in how much they actually constrain. Some pin down field types
and placement precisely; others — Table above among them — define one permissive
set of fields reused at every level, and will happily accept a document the
prose spec would call wrong. A clean schema check means nothing contradicts the
schema, not that the metadata is correct.

There is a gap in the core schemas too. STAC annotates its URL-bearing fields —
`href`, `providers[].url`, and the `stac_extensions` entries — with the JSON
Schema `iri` and `iri-reference` formats. The `ajv` validator underneath
`jsonvalidate` implements the `uri` family but not the internationalised `iri`
variants, so it skips those assertions: the fields are checked for type but not
for shape, and a malformed URL will pass. `ajv` warns once per occurrence when
it does this; `stacbuildr` filters those particular warnings out, since they say
nothing about the object being validated, and lets every other warning
through.

Here is the finished set of extensions on our scene:

```{r}
item@stac_extensions
```

## Where to go next

See `vignette("stac-catalog")` for assembling items into a catalog, writing it
to disk, and serving it through a STAC API.
