---
title: "Describing Data with STAC Extensions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Describing Data with STAC Extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The core STAC Item spec is deliberately small: an id, a geometry, a datetime,
and a bag of assets. Everything that makes a dataset *interpretable* — which
band is red, what the nodata value is, what pixel value 3 means, how to draw it
on a map — lives in **extensions**.

An extension is two things:

1. A schema URI added to the Item's `stac_extensions` array, which tells clients
   and validators which vocabulary is in play.
2. A set of prefixed fields (`eo:cloud_cover`, `raster:scale`,
   `classification:classes`, …) placed either in the Item's `properties`, on an
   individual asset, or — for anything describing a single band — inside the
   `bands` array shared by every band-level extension.

`stacbuildr` gives each supported extension an `add_*_extension()` function that
handles both halves. You never edit `stac_extensions` by hand.

This vignette covers the extensions the package supports, when to reach for
each, and the recurring decision of item-level versus asset-level placement.

## Setup

```{r}
library(stacbuildr)
```

We will build up a single Landsat-like scene, adding one extension at a time.

```{r}
item <- stac_item(
  id = "LC09_L2SP_044033_20230615",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-122.5, 39.5), c(-120.5, 39.5), c(-120.5, 40.5),
      c(-122.5, 40.5), c(-122.5, 39.5)
    ))
  ),
  bbox = c(-122.5, 39.5, -120.5, 40.5),
  datetime = "2023-06-15T18:45:37Z",
  properties = list(
    platform = "landsat-9",
    instruments = c("oli", "tirs")
  )
)

item <- item |>
  add_asset(
    key = "red",
    href = "https://example.com/LC09_B4.tif",
    type = "image/tiff; application=geotiff; profile=cloud-optimized",
    roles = c("data")
  ) |>
  add_asset(
    key = "nir",
    href = "https://example.com/LC09_B5.tif",
    type = "image/tiff; application=geotiff; profile=cloud-optimized",
    roles = c("data")
  )
```

## Item-level or asset-level?

Nearly every `add_*_extension()` function takes an `asset_key` argument, and the
choice it controls matters more than it first appears.

* **Omit `asset_key`** and the fields land in `item@properties`. This says the
  metadata describes the whole scene.
* **Pass `asset_key`** and the fields land on that asset. This says the metadata
  describes only that file.

The rule of thumb: if the value would differ between two assets in the same
item, it belongs on the asset. Band descriptions, data types, nodata values and
resolutions are per-file, so they go on assets. Scene-wide facts — cloud cover,
the DOI of the paper describing the dataset — go on the item.

Passing an `asset_key` that does not exist is an error, so add assets before
extensions:

```{r, error = TRUE}
add_eo_extension(
  item,
  bands = list(eo_band(name = "B6", common_name = "swir16")),
  asset_key = "swir"
)
```

One field resists this choice: `table:storage_options` describes how to *open a
particular file*, so it only ever makes sense on an asset. `add_table_extension()`
requires an `asset_key` when you supply it.

## Electro-Optical (EO)

The [EO extension](https://github.com/stac-extensions/eo) describes the spectral
properties of imagery: what wavelength each band samples, and how much of the
scene was obscured by cloud or snow.

`eo_band()` builds one band; `add_eo_extension()` attaches a list of them along
with the optional scene-level `cloud_cover` and `snow_cover` percentages (both
validated to 0–100).

```{r}
item <- item |>
  add_eo_extension(cloud_cover = 12.5, snow_cover = 0) |>
  add_eo_extension(
    bands = list(
      eo_band(
        name = "B4",
        common_name = "red",
        center_wavelength = 0.655,
        full_width_half_max = 0.037
      )
    ),
    asset_key = "red"
  ) |>
  add_eo_extension(
    bands = list(
      eo_band(
        name = "B5",
        common_name = "nir",
        center_wavelength = 0.865,
        full_width_half_max = 0.028
      )
    ),
    asset_key = "nir"
  )

item@properties$`eo:cloud_cover`
item@assets$red$bands
```

Note the pattern above: calling `add_eo_extension()` several times is fine. The
schema URI is only added once, and each call writes to a different place.

Bands land in the asset's `bands` array, not in an `eo:bands` array. Version
2.0.0 of the extension dropped `eo:bands` in favour of the single `bands` array
STAC 1.1 defines for every band-level extension, and moved its own fields
behind an `eo:` prefix — hence `eo:common_name` rather than `common_name` in
the output. The `eo_band()` arguments keep their short names.

`common_name` is checked against the STAC common band names (`"coastal"`,
`"blue"`, `"green"`, `"red"`, `"rededge"`, `"nir"`, `"swir16"`, `"lwir11"`,
and so on). It is what lets a client ask for "the red band" without knowing that
Landsat calls it B4 and Sentinel-2 calls it B04.

### Sensor presets

Typing out band definitions for a well-known sensor is tedious and easy to get
wrong, so the package ships presets:

```{r}
bands <- landsat_oli_bands()
length(bands)
bands[[4]][c("name", "eo:common_name", "eo:center_wavelength")]
```

The others are `sentinel2_msi_bands()`, `worldview3_bands()`, `skysat_bands()`
and `planetscope_bands()`. `landsat_oli_bands()` takes
`include_thermal = TRUE` to append the TIRS bands.

## Raster

Where EO describes the *physics* of a band, the
[Raster extension](https://github.com/stac-extensions/raster) describes the
*pixels*: data type, nodata value, ground sample distance, and the scale/offset
needed to turn stored integers into physical units.

`raster_band()` is an S7 object rather than a plain list, so typos in
`data_type` are caught as they are written.

```{r}
red_band <- raster_band(
  nodata = 0,
  data_type = "uint16",
  spatial_resolution = 30,
  scale = 0.0000275,
  offset = -0.2,
  unit = "reflectance",
  statistics = raster_statistics(
    minimum = 1,
    maximum = 65455,
    mean = 9821,
    stddev = 3110,
    valid_percent = 98.7
  )
)

item <- add_raster_extension(item, bands = list(red_band), asset_key = "red")

item@assets$red$bands[[1]]$`raster:scale`
```

Raster v2.0.0 made the same move as EO: the band objects go in the shared
`bands` array, `nodata`, `data_type`, `unit` and `statistics` stay unprefixed as
STAC Common Metadata, and the extension's own fields gain a `raster:` prefix on
the way out. The `raster_band()` arguments are unchanged.

`raster_statistics()` and `raster_histogram()` build the optional `statistics`
and `histogram` sub-objects. Statistics are worth including: they let a client
pick sensible display limits without reading the file.

### Deriving bands from a file

Rather than transcribing metadata that GDAL already knows,
`band_from_file()` reads it straight off the raster. With
`calculate_statistics = TRUE` it also samples the file to populate the
statistics block.

```{r}
tif <- system.file("ex/elev.tif", package = "terra")

if (requireNamespace("terra", quietly = TRUE) && nzchar(tif)) {
  bands <- band_from_file(tif, calculate_statistics = TRUE)
  str(bands[[1]], max.level = 1)
}
```

`band_from_file()` already returns a *list* of bands, one per layer. Pass it
straight through — wrapping it again in `list()` is the most common mistake here,
and the function raises a specific error when it detects the double-wrapping.

```{r, error = TRUE}
add_raster_extension(item, bands = list(band_from_file(tif)))
```

`bands_from_terra()` does the same job for a `SpatRaster` you already have in
memory.

### EO and Raster together

The two extensions describe the same bands from different angles, and since
both now write to the one `bands` array, a single band object carries both
vocabularies. Common fields (`name`, `description`, `nodata`, `data_type`) are
unprefixed; extension-specific ones are prefixed.

Describe the bands with each extension in turn and the fields are merged, band
by band, so long as both calls pass the same number of bands:

```{r}
combined_item <- item |>
  add_eo_extension(
    bands = list(eo_band(name = "B4", common_name = "red",
                         center_wavelength = 0.655)),
    asset_key = "red"
  ) |>
  add_raster_extension(
    bands = list(raster_band(nodata = 0, data_type = "uint16",
                             spatial_resolution = 30, scale = 0.0000275)),
    asset_key = "red"
  )

names(combined_item@assets$red$bands[[1]])
```

A list of a different length is taken to describe different bands and replaces
what was there, so build one band per band. The other route is to write a single
band object yourself, passing the raster fields to `eo_band()` through `...`:

```{r}
combined <- eo_band(
  name = "B4",
  common_name = "red",
  center_wavelength = 0.655,
  nodata = 0,
  data_type = "uint16",
  "raster:spatial_resolution" = 30,
  "raster:scale" = 0.0000275
)

names(combined)
```

## Classification

The [Classification extension](https://github.com/stac-extensions/classification)
gives meaning to the integer values in a thematic raster. It has two mutually
exclusive modes, and `add_classification_extension()` errors if you supply both.

**Classes** map a pixel value to a name — land cover maps, crop type layers:

```{r}
lc_item <- stac_item(
  id = "nlcd-2021",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-122.5, 39.5), c(-120.5, 39.5), c(-120.5, 40.5),
      c(-122.5, 40.5), c(-122.5, 39.5)
    ))
  ),
  bbox = c(-122.5, 39.5, -120.5, 40.5),
  datetime = "2021-01-01T00:00:00Z"
) |>
  add_asset(
    key = "landcover",
    href = "https://example.com/nlcd-2021.tif",
    type = "image/tiff; application=geotiff",
    roles = c("data")
  )

classes <- list(
  classification_class(value = 0, name = "nodata", nodata = TRUE),
  classification_class(value = 11, name = "open-water",
                       title = "Open Water", color_hint = "466B9F"),
  classification_class(value = 42, name = "evergreen-forest",
                       title = "Evergreen Forest", color_hint = "1C5F2C"),
  classification_class(value = 82, name = "cultivated-crops",
                       title = "Cultivated Crops", color_hint = "AB6C28")
)

lc_item <- add_classification_extension(
  lc_item,
  classes = classes,
  asset_key = "landcover"
)
```

Two constraints catch mistakes early: `name` must be a slug (letters, digits,
hyphens, underscores — no spaces, which is why the titles above carry the
human-readable form), and `color_hint` must be exactly six upper-case hex digits
with no leading `#`.

```{r, error = TRUE}
classification_class(value = 11, name = "open water")
```

**Bitfields** describe QA bands where several flags are packed into one integer.
Each `classification_bitfield()` names a bit range by `offset` and `length` and
carries its own classes:

```{r}
qa_item <- add_classification_extension(
  lc_item,
  bitfields = list(
    classification_bitfield(
      offset = 0, length = 1, name = "fill",
      description = "Image or fill data",
      classes = list(
        classification_class(value = 0, name = "not-fill"),
        classification_class(value = 1, name = "fill")
      )
    ),
    classification_bitfield(
      offset = 3, length = 1, name = "cloud",
      description = "Cloud detected",
      classes = list(
        classification_class(value = 0, name = "not-cloud"),
        classification_class(value = 1, name = "cloud")
      )
    )
  )
)

vapply(qa_item@properties$`classification:bitfields`, `[[`, character(1), "name")
```

## Render

The [Render extension](https://github.com/stac-extensions/render) records how a
dataset is *meant* to be displayed — which assets to combine, how to stretch
them, which colormap to use. Dynamic tilers such as titiler read these
directly, so a viewer can produce the intended visualization without anyone
hardcoding parameters.

`renders` is a **named** list; the names become the keys under `renders` and are
what a client shows in a layer picker.

```{r}
item <- add_render_extension(
  item,
  renders = list(
    natural_color = render_object(
      assets = c("red", "nir"),
      title = "Natural Color",
      rescale = list(c(0, 10000), c(0, 10000)),
      resampling = "bilinear"
    ),
    ndvi = render_object(
      assets = c("nir", "red"),
      title = "NDVI",
      expression = "(nir-red)/(nir+red)",
      rescale = list(c(-1, 1)),
      colormap_name = "ylgn",
      minmax_zoom = c(6, 14)
    )
  )
)

names(item@properties$renders)
```

`rescale` is a list of two-element numeric vectors, one per band being
stretched — a flat `c(0, 10000)` is rejected. Render is the one extension here
that also accepts a `stac_collection()`, which is usually where it belongs: the
display recipe rarely varies scene to scene.

```{r}
collection <- stac_collection(
  id = "landsat-c2-l2",
  description = "Landsat Collection 2 Level-2 surface reflectance",
  license = "CC0-1.0",
  extent = stac_extent(
    spatial_bbox = list(c(-180, -90, 180, 90)),
    temporal_interval = list(list("2013-02-11T00:00:00Z", NULL))
  )
) |>
  add_render_extension(renders = list(
    natural_color = render_object(
      assets = c("red", "nir"),
      title = "Natural Color",
      rescale = list(c(0, 10000), c(0, 10000))
    )
  ))
```

## Scientific citation

The [Scientific extension](https://github.com/stac-extensions/scientific) records
how a dataset should be cited, and which publications describe it. Fields go in
the item properties — there is no `asset_key`, since a citation applies to the
dataset as a whole.

```{r}
item <- add_scientific_extension(
  item,
  doi = "10.5066/P9OGBGM6",
  citation = "U.S. Geological Survey (2023). Landsat 9 Collection 2 Level-2.",
  publications = list(
    scientific_publication(
      doi = "10.1016/j.rse.2020.111968",
      citation = "Masek, J. et al. (2020). Landsat 9: Empowering open science."
    )
  )
)

item@properties$`sci:doi`
```

`doi` must be a bare DOI name, not a URL — clients build the resolver link
themselves:

```{r, error = TRUE}
add_scientific_extension(item, doi = "https://doi.org/10.5066/P9OGBGM6")
```

## Table

The [Table extension](https://github.com/stac-extensions/table) describes tabular
assets — GeoParquet, CSV, Arrow — so a client knows the schema before
downloading. Columns are built with `table_column()`; `primary_geometry` names
the geometry column, and `storage_options` carries the arguments a reader needs
to open the file.

`asset_key` behaves as it does elsewhere: omit it and the columns describe the
item, supply it and they describe that one asset. The default — omitting it — places
 `table:columns`, `table:primary_geometry` and `table:row_count` under "Item Properties and
Collection Fields". That is the right choice for the common case of an item
wrapping a single table, so the example below omits it for those fields.

`storage_options` is the exception, and the one field `asset_key` is mandatory
for. The README gives it its own "Asset Object fields" section: it holds
fsspec-style arguments for opening one specific file, so it has no meaning at
item level.

```{r}
parcels <- stac_item(
  id = "parcels-2023",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-122.5, 39.5), c(-120.5, 39.5), c(-120.5, 40.5),
      c(-122.5, 40.5), c(-122.5, 39.5)
    ))
  ),
  bbox = c(-122.5, 39.5, -120.5, 40.5),
  datetime = "2023-01-01T00:00:00Z"
) |>
  add_asset(
    key = "data",
    href = "https://example.com/parcels.parquet",
    type = "application/x-parquet",
    roles = c("data")
  ) |>
  add_table_extension(
    columns = list(
      table_column(name = "geometry", type = "binary",
                   description = "Parcel boundary (WKB)"),
      table_column(name = "parcel_id", type = "int64"),
      table_column(name = "area_ha", type = "double"),
      table_column(name = "zoning", type = "string")
    ),
    primary_geometry = "geometry",
    row_count = 148302
  ) |>
  add_table_extension(
    storage_options = list(account_name = "parcelsdata"),
    asset_key = "data"
  )

vapply(parcels@properties$`table:columns`, `[[`, character(1), "name")
names(parcels@assets$data$`table:storage_options`)
```

When one item bundles several tables with different schemas, pass `asset_key`
and each asset carries its own column list:

```{r}
multi <- stac_item(
  id = "census-2021",
  geometry = list(type = "Point", coordinates = c(-121.5, 40)),
  bbox = c(-121.5, 40, -121.5, 40),
  datetime = "2021-01-01T00:00:00Z"
) |>
  add_asset(key = "households", href = "https://example.com/hh.parquet",
            type = "application/x-parquet") |>
  add_asset(key = "persons", href = "https://example.com/persons.parquet",
            type = "application/x-parquet") |>
  add_table_extension(
    columns = list(table_column(name = "household_id", type = "int64"),
                   table_column(name = "tenure", type = "string")),
    row_count = 24500,
    asset_key = "households"
  ) |>
  add_table_extension(
    columns = list(table_column(name = "person_id", type = "int64"),
                   table_column(name = "age", type = "int32")),
    row_count = 61200,
    asset_key = "persons"
  )

vapply(multi@assets$persons$`table:columns`, `[[`, character(1), "name")
```

One caveat on how much the spec backs any of this. The item/asset split is
stated only in the README's field tables; the extension's JSON schema reuses a
single permissive `fields` definition for properties, assets and `item_assets`
alike, so a validator will not object wherever you put the columns — and it does
not mention `table:storage_options` at all. Table is a *Pilot*-maturity
extension. Treat the placement above as the documented convention rather than
something a schema check enforces.

## Vector

The [Vector extension](https://github.com/stac-extensions/vector) records the
cartographic properties of a vector dataset: which geometry types it contains,
the minimum mapping unit and width used when it was generalized, and the scale
it was digitized for. These tell a user whether the data is fit for their
purpose — a layer captured at 1:50,000 should not be used for parcel-level work.

```{r}
parcels <- add_vector_extension(
  parcels,
  geometry_types = c("Polygon", "MultiPolygon"),
  mmu = 100,
  mmw = 5,
  reference_scale = 50000
)

parcels@properties[c("vector:geometry_types", "vector:reference_scale")]
```

`geometry_types` is checked against the seven GeoJSON geometry type names, so
lower-case or pluralized spellings fail rather than reaching a validator later.

Table and Vector are complementary and frequently appear together: Table
describes the columns, Vector describes the geometry.

## Datacube

The [Datacube extension](https://github.com/stac-extensions/datacube) describes
multidimensional arrays — NetCDF, Zarr, HDF — where the interesting structure is
the axes rather than the bands. It has two halves: `dimensions` names the axes,
and `variables` says which arrays exist over them.

Both are **named** lists, and the names in a variable's `dimensions` argument
must refer to dimensions you defined.

```{r}
cube <- stac_item(
  id = "era5-2023-06",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-180, -90), c(180, -90), c(180, 90), c(-180, 90), c(-180, -90)
    ))
  ),
  bbox = c(-180, -90, 180, 90),
  datetime = "2023-06-01T00:00:00Z"
) |>
  add_datacube_extension(
    dimensions = list(
      lon = cube_dimension(
        type = "spatial", axis = "x",
        extent = c(-180, 180), step = 0.25,
        reference_system = 4326
      ),
      lat = cube_dimension(
        type = "spatial", axis = "y",
        extent = c(-90, 90), step = 0.25,
        reference_system = 4326
      ),
      time = cube_dimension(
        type = "temporal",
        extent = c("2023-06-01T00:00:00Z", "2023-06-30T23:00:00Z"),
        step = "PT1H"
      ),
      level = cube_dimension(
        type = "vertical",
        values = c(1000, 850, 500, 250),
        unit = "hPa"
      )
    ),
    variables = list(
      temperature = cube_variable(
        type = "data",
        dimensions = c("lon", "lat", "time", "level"),
        unit = "K",
        data_type = "float32",
        description = "Air temperature"
      ),
      land_mask = cube_variable(
        type = "auxiliary",
        dimensions = c("lon", "lat"),
        data_type = "uint8"
      )
    )
  )

names(cube@properties$`cube:dimensions`)
```

A few points that are easy to get wrong:

* `type = "spatial"` requires an `axis` of `"x"`, `"y"` or `"z"`, and horizontal
  axes require a two-element `extent`.
* Use `extent` for a continuous range and `values` for an enumerated set — the
  pressure levels above are enumerated because they are not evenly spaced.
* `step` is a number for numeric axes and an ISO 8601 duration for temporal ones
  (`"PT1H"` is one hour). Set it to `NULL` for irregular spacing.
* Variables are `"data"` (the payload) or `"auxiliary"` (supporting arrays such
  as masks or coordinate lookups).

## Point Cloud

The [Point Cloud extension](https://github.com/stac-extensions/pointcloud)
describes datasets of discrete points — LiDAR most often, but also radar, sonar,
and point clouds matched from overlapping imagery. Two fields are required,
`pc:count` and `pc:type`, and three more are optional: `pc:density`,
`pc:schemas` (the dimensions each point carries) and `pc:statistics`
(per-channel summaries).

```{r}
cloud <- stac_item(
  id = "als-tile-042",
  geometry = list(
    type = "Polygon",
    coordinates = list(list(
      c(-78.644, 45.289), c(-78.641, 45.289),
      c(-78.641, 45.291), c(-78.644, 45.291), c(-78.644, 45.289)
    ))
  ),
  bbox = c(-78.644, 45.289, -78.641, 45.291),
  datetime = "2023-06-15T00:00:00Z"
)

cloud <- add_pointcloud_extension(
  cloud,
  count   = 81590,
  type    = "lidar",
  density = 1.54,
  schemas = list(
    pc_schema("X", size = 8, type = "floating"),
    pc_schema("Y", size = 8, type = "floating"),
    pc_schema("Z", size = 8, type = "floating"),
    pc_schema("Intensity", size = 2, type = "unsigned")
  ),
  statistics = list(
    pc_statistic("Z", position = 2, minimum = 0, maximum = 29.97)
  )
)

cloud@properties[c("pc:count", "pc:type", "pc:density")]
```

Two details are worth knowing before you write these by hand.

`pc:type` is **not** an enumeration, unlike most controlled STAC fields. The
specification suggests `"lidar"`, `"eopc"`, `"radar"`, `"sonar"` and `"other"`,
but its schema accepts any non-empty string. So an unrecognised value warns and
is kept, rather than failing the way an invalid `vector:geometry_types` would:

```{r}
photogrammetric <- add_pointcloud_extension(
  cloud, count = 100, type = "structure-from-motion"
)
```

Sizes in `pc:schemas` are in **whole bytes**, which the LAS format does not
oblige. `ReturnNumber`, `NumberOfReturns`, `ScanDirectionFlag` and the
classification flags are bit-packed into shared bytes in the file itself. The
convention, set by PDAL and followed here, is to describe each as the unpacked
one-byte dimension a reader materialises — the file layout is not what the field
is for.

### From a LAS or LAZ file

`item_from_lidr()` fills all of this in from a point cloud's header, via the
[lidR](https://r-lidar.github.io/lidRbook/) package. Everything above comes out
of the public header block, so no points are read and the cost does not grow
with the size of the file:

```{r, eval = requireNamespace("lidR", quietly = TRUE)}
f <- system.file("extdata", "Megaplot.laz", package = "lidR")

tile <- item_from_lidr(f, datetime = "2023-06-15T00:00:00Z")

tile@properties[c("pc:count", "pc:density", "proj:code")]
```

The dimension schema is derived from the file's point data record format and
from any Extra Bytes record it declares:

```{r, eval = requireNamespace("lidR", quietly = TRUE)}
vapply(tile@properties$`pc:schemas`, function(s) s$name, character(1))
```

Statistics are the one thing the header cannot fully supply. It carries the X,
Y and Z bounds, so those channels get `minimum`, `maximum` and `count` for free;
averages, standard deviations and any other channel need a full read, which
`calculate_statistics = TRUE` opts into.

For a tiled collection, `items_from_lascatalog()` takes a `lidR` `LAScatalog`
(or a directory, or a vector of paths) and returns one item per file, skipping
any tile it cannot read rather than abandoning the run. Pass the result to
`extent_from_items()` to derive the collection extent.

A note on media types: `get_media_type()` recognises `.las` and `.laz`, and maps
the `.copc.laz` convention to `application/vnd.laszip+copc`. Only the COPC type
is listed in the STAC best practices — it is the cloud-optimized point cloud
format, and the one worth reaching for if the data will be read over HTTP.

## Projection

STAC requires an Item's `geometry` and `bbox` to be WGS84 longitude/latitude
whatever projection the data is stored in. The Projection extension carries the
native CRS alongside it, so a client can locate a pixel or read a window without
opening the file.

Most of the time it is applied for you. `item_from_terra()` populates
`proj:code`, `proj:wkt2`, `proj:shape`, `proj:bbox` and `proj:transform` from
the raster whenever its CRS is not WGS84; `item_from_lidr()` does the same for a
point cloud, where `proj:bbox` is three-dimensional because the LAS header
carries Z bounds as well.

`add_projection_extension()` does it by hand, for an item built from something
other than a raster or point cloud:

```{r}
proj_item <- stac_item(
  id = "utm-scene",
  geometry = list(type = "Point", coordinates = c(-113.5, 51.0)),
  bbox = c(-113.5, 51.0, -113.5, 51.0),
  datetime = "2024-06-01T00:00:00Z"
) |>
  add_projection_extension(
    code = "EPSG:32612",
    shape = c(5558, 9559),
    transform = c(30, 0, 712710, 0, -30, 5654790),
    bbox = c(712710, 5487090, 999480, 5654790)
  )

proj_item@properties[c("proj:code", "proj:shape")]
```

`proj:code` is an `AUTHORITY:CODE` string — it replaced the bare-integer
`proj:epsg` in v2.0.0 of the extension, so `"32612"` on its own is rejected and
`"EPSG:32612"` is required. That also lets authorities other than EPSG be
recorded, such as `"OGC:CRS84"`.

Like the other extensions, `asset_key` moves the fields onto a single asset.
This is what a Sentinel-2 scene needs: its 10 m, 20 m and 60 m bands share a
CRS but differ in `proj:shape` and `proj:transform`, so the CRS goes on the item
and the grid on each asset.

An item already in EPSG:4326 needs none of this — the geometry and bbox say
everything.

## Checking the result

Two levels of checking are available. `validate_stac()` runs the package's own
structural rules and needs no extra dependencies:

```{r}
result <- validate_stac(item)
result$valid
```

`validate_stac_schema()` goes further, fetching the official JSON schemas and
validating the serialized item against the core spec *and* against every schema
listed in `stac_extensions`. It requires the `jsonvalidate` package and network
access, which is why it is not run here:

```{r, eval = FALSE}
validate_stac_schema(item, validate_extensions = TRUE)
```

The second form is worth running once over a representative item before
publishing a catalog. Temper your expectations of it, though: extension schemas
vary a great deal in how much they actually constrain. Some pin down field types
and placement precisely; others — Table above among them — define one permissive
set of fields reused at every level, and will happily accept a document the
prose spec would call wrong. A clean schema check means nothing contradicts the
schema, not that the metadata is correct.

There is a gap in the core schemas too. STAC annotates its URL-bearing fields —
`href`, `providers[].url`, and the `stac_extensions` entries — with the JSON
Schema `iri` and `iri-reference` formats. The `ajv` validator underneath
`jsonvalidate` implements the `uri` family but not the internationalised `iri`
variants, so it skips those assertions: the fields are checked for type but not
for shape, and a malformed URL will pass. `ajv` warns once per occurrence when
it does this; `stacbuildr` filters those particular warnings out, since they say
nothing about the object being validated, and lets every other warning
through.

Here is the finished set of extensions on our scene:

```{r}
item@stac_extensions
```

## Where to go next

See `vignette("stac-catalog")` for assembling items into a catalog, writing it
to disk, and serving it through a STAC API.
