---
title: "Building a STAC Catalog"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a STAC Catalog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to build a STAC catalog with `stacbuildr`, covering:

1. Creating a catalog, collection, and item from a raster file
2. Writing a static catalog to disk
3. Loading into a PostgreSQL database and serving a STAC API *(requires infrastructure)*

## Setup

```{r}
library(stacbuildr)
library(terra)
```

## Create a synthetic DEM

For this example we create a small synthetic Digital Elevation Model and write
it to a temporary file as a Cloud Optimized GeoTIFF. In practice, replace
`tif_path` with the path to your own raster file.

`filetype = "COG"` selects GDAL's COG driver, which writes a tiled,
internally-ordered GeoTIFF that clients can read over HTTP range requests
without downloading the whole file. This example raster is only 10×10, so it
fits in a single block and GDAL builds no overview levels — a real dataset large
enough to need them gets those automatically.

```{r}
set.seed(42)
dem <- terra::rast(nrows = 10, ncols = 10, xmin = -120, xmax = -119,
                   ymin = 48, ymax = 49, crs = "EPSG:4326")
terra::values(dem) <- runif(terra::ncell(dem), 100, 3000)
names(dem) <- "elevation"

data_dir <- file.path(tempdir(), "data")
dir.create(data_dir, showWarnings = FALSE)

tif_path <- file.path(data_dir, "dem.tif")
terra::writeRaster(
  dem,
  tif_path,
  filetype = "COG",
  gdal = c("COMPRESS=DEFLATE", "BLOCKSIZE=256"),
  overwrite = TRUE
)

r <- terra::rast(tif_path)
```

GDAL records the COG layout in the file's structural metadata, which is what
justifies the `profile=cloud-optimized` media type used on the asset below:

```{r}
grepl(
  "LAYOUT=COG",
  sf::gdal_utils("info", source = tif_path, quiet = TRUE),
  fixed = TRUE
)
```

## Create a STAC Item from the raster

`item_from_terra()` extracts the spatial extent, CRS, and geometry from the
`SpatRaster` object and returns a `stac_item`. We then attach the raster file
as an asset.

```{r}
item <- r |>
  item_from_terra(
    id = "dem-001",
    datetime = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ")
  ) |>
  add_asset(
    key = "dem",
    href = tif_path,
    title = "Digital Elevation Model",
    description = "Synthetic DEM for demonstration",
    type = "image/tiff; application=geotiff; profile=cloud-optimized",
    roles = "data"
  )

item
```

## Create a Collection and Catalog

```{r}
collection <- stac_collection(
  id = "terrain",
  description = "Terrain datasets",
  title = "Terrain Collection",
  license = "CC-BY-4.0",
  extent = stac_extent(
    spatial_bbox = list(c(terra::xmin(r), terra::ymin(r), terra::xmax(r), terra::ymax(r))),
    temporal_interval = list(list("2020-01-01T00:00:00Z", NULL))
  )
)

collection
```

```{r}
catalog <- stac_catalog(
  id = "my-catalog",
  description = "Example STAC catalog",
  title = "My Catalog"
)

collection <- add_item(collection, item)
catalog <- add_child(catalog, collection)

print(catalog, expand = TRUE)
```

## Write a static catalog to disk

`write_stac()` recursively writes the catalog, collection, and item JSON files.
The `catalog_type` argument controls how links and asset hrefs are written:

- `"self-contained"` — links between catalog/collection/item files, and asset
  hrefs given as absolute local paths, are rewritten relative to the file that
  contains them. Use for sharing or archiving.
- `"relative"` — currently identical to `"self-contained"`.
- `"absolute"` — all links use full URLs built from `base_url`. Use when
  publishing to a web server so remote clients can crawl the catalog over HTTP.
  Assets should also be URLs (S3, HTTPS) in this case.

One thing to be aware of: `write_stac()` writes JSON and nothing else. It does
not copy asset files into the catalog directory. Our DEM lives in `data_dir`,
which is a sibling of the catalog rather than inside it, so the written item
gets an href like `../../../data/dem.tif` — a valid relative path, but one that
points outside the tree.

That is narrower than the STAC best-practices meaning of "self-contained", where
every referenced file sits inside the catalog directory so the folder can be
zipped and moved as a unit. If you need that, copy the assets under
`catalog_dir` yourself before calling `write_stac()`.

We use `"self-contained"` here:

```{r}
catalog_dir <- file.path(tempdir(), "catalog")

write_stac(
  catalog,
  catalog_dir,
  catalog_type = "self-contained",
  overwrite = TRUE
)
```

The resulting directory structure looks like:

```
catalog/
  catalog.json
  terrain/
    collection.json
    dem-001/
      dem-001.json
```

We can read it back:

```{r}
cat_read <- read_stac(file.path(catalog_dir, "catalog.json"))
coll_read <- read_stac(file.path(catalog_dir, "terrain", "collection.json"))
item_read <- read_stac(file.path(
  catalog_dir,
  "terrain",
  "dem-001",
  "dem-001.json"
))

item_read
```

## Database-backed STAC API

The following sections require a running PostgreSQL database and (optionally)
a Posit Connect server to deploy the API. They are shown for reference but are
not evaluated when the vignette is built.

### Set up the database

`stac_db_setup()` creates tables and indexes inside an existing database; it
cannot create the database itself, because PostgreSQL only accepts
`CREATE DATABASE` from a connection to a *different* database. Create it once
up front, either from the shell:

```sh
createdb stac
```

or from R, by connecting to the default `postgres` maintenance database:

```{r, eval = FALSE}
library(DBI)

admin <- dbConnect(
  RPostgres::Postgres(),
  host = "localhost",
  port = 5432,
  dbname = "postgres",
  user = Sys.getenv("PG_USER"),
  password = Sys.getenv("PG_PASSWORD")
)

if (nrow(dbGetQuery(
  admin,
  "SELECT 1 FROM pg_database WHERE datname = 'stac'"
)) == 0) {
  dbExecute(admin, "CREATE DATABASE stac")
}

dbDisconnect(admin)
```

The role also needs permission to run `CREATE EXTENSION postgis`, which
`stac_db_setup()` issues on first use. A superuser connection, or a database
created from the `template_postgis` template, satisfies this.

Now connect to the `stac` database and create the schema:

```{r, eval = FALSE}
con <- dbConnect(
  RPostgres::Postgres(),
  host = "localhost",
  port = 5432,
  dbname = "stac",
  user = Sys.getenv("PG_USER"),
  password = Sys.getenv("PG_PASSWORD")
)

stac_db_setup(con)
```

### Serve the asset over HTTP

The item built above has `href = tif_path`, a local filesystem path. That works
for the static catalog, where `write_stac()` rewrites the href to a path
relative to the item JSON and a local reader resolves it on disk. It does
**not** work for the API: a client such as QGIS resolves a non-URL href against the API
base URL, producing a nonsense address like
`http://127.0.0.1:3485/var/folders/.../dem.tif`, and the request 404s.

The STAC API serves JSON only — it never serves asset bytes. So assets must
already be fetchable over HTTP from somewhere else. For local development, run
a static file server rooted at `data_dir`, the directory the DEM was written to.
Print the exact command and paste it into a terminal, leaving it running:

```{r}
cat(sprintf('npx http-server "%s" -p 8000 --cors\n', data_dir))
```

`npx serve "<data_dir>" -l 8000 --cors` works equally well.

Three details are easy to get wrong here:

- **Serve `data_dir`, not `$TMPDIR`.** R creates a per-session
  `Rtmp<XXXXXX>` directory beneath `$TMPDIR`, so a server rooted at `$TMPDIR`
  leaves the raster one path segment deeper than the href says, and every
  request 404s.
- **Use a server that honours `Range`.** GDAL reads remote rasters through
  `/vsicurl/`, which issues HTTP range requests to fetch only the tiles and
  overviews it needs. Python's built-in `http.server` ignores `Range` — it
  answers `200` with the full body and advertises no `Accept-Ranges` — forcing
  the whole file over the wire on every read, which is ruinous for a
  cloud-optimized GeoTIFF. Both `npx` servers above return `206 Partial
  Content` correctly.
- **Keep the R session alive.** R deletes `tempdir()` on exit, taking the
  raster with it and leaving the stored hrefs pointing at nothing. For anything
  beyond this walkthrough, put assets in a persistent directory.

`--cors` is what lets browser-based clients such as STAC Browser read the
assets; QGIS does not need it, but it costs nothing.

Then replace the asset with one pointing at that server. Re-using the same key
overwrites the existing asset:

```{r, eval = FALSE}
item <- add_asset(
  item,
  key = "dem",
  href = paste0("http://127.0.0.1:8000/", basename(tif_path)),
  title = "Digital Elevation Model",
  description = "Synthetic DEM for demonstration",
  type = "image/tiff; application=geotiff; profile=cloud-optimized",
  roles = "data"
)
```

Confirm the URL resolves before inserting — `curl -I <href>` should return
`200 OK`. In production, use blob storage URLs (S3, Azure Blob, GCS) so clients
fetch assets directly without going through the API or a local file server.

### Insert a collection and item

Insert *after* fixing the href: `stac_db_insert_item()` stores the item as-is,
so an href inserted wrong stays wrong until you re-insert.

```{r, eval = FALSE}
stac_db_insert_collection(con, collection)

item@collection <- "terrain"
stac_db_insert_item(con, item)
```

### Run the STAC API locally

`stac_api_router()` returns a plumber router pre-wired with all STAC endpoints.
For local development, pass `require_auth = FALSE` to skip authentication.

```{r, eval = FALSE}
library(plumber)

pr <- stac_api_router(
  con,
  base_url = "http://127.0.0.1:3485"
)

pr_run(pr, port = 3485)
```

### Deploying to Posit Connect

When deploying to Posit Connect with multiple concurrent users, replace the
single DBI connection with a connection pool from the `pool` package. A pool
manages multiple connections and hands them out to simultaneous requests without
contention. Store credentials in environment variables rather than hardcoding
them — Posit Connect lets you set these per-deployment under the Vars tab.

```{r, eval = FALSE}
library(pool)
library(plumber)

pool <- dbPool(
  RPostgres::Postgres(),
  host = Sys.getenv("PG_HOST"),
  dbname = Sys.getenv("PG_DBNAME"),
  user = Sys.getenv("PG_USER"),
  password = Sys.getenv("PG_PASSWORD")
)

onStop(function() poolClose(pool))

pr <- stac_api_router(
  pool,
  base_url = Sys.getenv("STAC_BASE_URL") # e.g. "https://connect.example.com/stac"
)

pr_run(pr)
```
