---
title: "Building a STAC Catalog"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building a STAC Catalog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to build a STAC catalog with `stacbuildr`, covering:

1. Creating a catalog, collection, and item from a raster file
2. Writing a static catalog to disk
3. Loading into a PostgreSQL database and serving a STAC API *(requires infrastructure)*

## Setup

```{r}
library(stacbuildr)
library(terra)
```

## Create a synthetic DEM

For this example we create a small synthetic Digital Elevation Model and write
it to a temporary file as a Cloud Optimized GeoTIFF. In practice, replace
`tif_path` with the path to your own raster file.

`filetype = "COG"` selects GDAL's COG driver, which writes a tiled,
internally-ordered GeoTIFF that clients can read over HTTP range requests
without downloading the whole file. This example raster is only 10×10, so it
fits in a single block and GDAL builds no overview levels — a real dataset large
enough to need them gets those automatically.

```{r}
set.seed(42)
dem <- terra::rast(nrows = 10, ncols = 10, xmin = -120, xmax = -119,
                   ymin = 48, ymax = 49, crs = "EPSG:4326")
terra::values(dem) <- runif(terra::ncell(dem), 100, 3000)
names(dem) <- "elevation"

data_dir <- file.path(tempdir(), "data")
dir.create(data_dir, showWarnings = FALSE)

tif_path <- file.path(data_dir, "dem.tif")
terra::writeRaster(
  dem,
  tif_path,
  filetype = "COG",
  gdal = c("COMPRESS=DEFLATE", "BLOCKSIZE=256"),
  overwrite = TRUE
)

r <- terra::rast(tif_path)
```

GDAL records the COG layout in the file's structural metadata, which is what
justifies the `profile=cloud-optimized` media type used on the asset below:

```{r}
grepl(
  "LAYOUT=COG",
  sf::gdal_utils("info", source = tif_path, quiet = TRUE),
  fixed = TRUE
)
```

## Create a STAC Item from the raster

`item_from_terra()` extracts the spatial extent, CRS, and geometry from the
`SpatRaster` object and returns a `stac_item`. We then attach the raster file
as an asset.

```{r}
item <- r |>
  item_from_terra(
    id = "dem-001",
    datetime = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ")
  ) |>
  add_asset(
    key = "dem",
    href = tif_path,
    title = "Digital Elevation Model",
    description = "Synthetic DEM for demonstration",
    type = "image/tiff; application=geotiff; profile=cloud-optimized",
    roles = "data"
  )

item
```

## Create a Collection and Catalog

```{r}
collection <- stac_collection(
  id = "terrain",
  description = "Terrain datasets",
  title = "Terrain Collection",
  license = "CC-BY-4.0",
  extent = stac_extent(
    spatial_bbox = list(c(terra::xmin(r), terra::ymin(r), terra::xmax(r), terra::ymax(r))),
    temporal_interval = list(list("2020-01-01T00:00:00Z", NULL))
  )
)

collection
```

```{r}
catalog <- stac_catalog(
  id = "my-catalog",
  description = "Example STAC catalog",
  title = "My Catalog"
)

collection <- add_item(collection, item)
catalog <- add_child(catalog, collection)

print(catalog, expand = TRUE)
```

## Write a static catalog to disk

`write_stac()` recursively writes the catalog, collection, and item JSON files.
The `catalog_type` argument controls how links and asset hrefs are written. The
three types correspond to the link layouts defined in
[Use of links](https://github.com/radiantearth/stac-spec/blob/master/best-practices.md#use-of-links)
in the STAC best-practices document, and to PySTAC's `SELF_CONTAINED`,
`RELATIVE_PUBLISHED` and `ABSOLUTE_PUBLISHED`:

- [`"self-contained"`](https://github.com/radiantearth/stac-spec/blob/master/best-practices.md#self-contained-catalogs) — links between
  catalog/collection/item files, and asset hrefs given as absolute local paths,
  are rewritten relative to the file that contains them. No object gets a
  `self` link, because a `self` link has to be absolute and that would tie the
  catalog to one location. Use for sharing or archiving.
- [`"relative"`](https://github.com/radiantearth/stac-spec/blob/master/best-practices.md#relative-published-catalog) — a self-contained catalog plus
  a single absolute `self` link on the root, recording where the catalog is
  published. Requires `base_url`. Use when the catalog is online but should
  still work after being downloaded.
- [`"absolute"`](https://github.com/radiantearth/stac-spec/blob/master/best-practices.md#absolute-published-catalog) — all links use full URLs built
  from `base_url`, and every object gets a `self` link. Use when publishing to a
  web server so remote clients can crawl the catalog over HTTP. Asset hrefs that
  are already URLs (S3, HTTPS) are kept; relative ones are resolved against the
  item's URL.

One thing to be aware of: `write_stac()` writes JSON and nothing else. It does
not copy asset files into the catalog directory. Our DEM lives in `data_dir`,
which is a sibling of the catalog rather than inside it, so the written item
gets an href like `../../../data/dem.tif` — a valid relative path, but one that
points outside the tree.

That is narrower than a
[self-contained catalog with assets](https://github.com/radiantearth/stac-spec/blob/master/best-practices.md#self-contained-with-assets), where every
referenced file sits inside the catalog directory so the folder can be zipped
and moved as a unit — what we write here is
[metadata only](https://github.com/radiantearth/stac-spec/blob/master/best-practices.md#self-contained-metadata-only). If you need the former, copy
the assets under `catalog_dir` yourself before calling `write_stac()`.

We use `"self-contained"` here:

```{r}
catalog_dir <- file.path(tempdir(), "catalog")

write_stac(
  catalog,
  catalog_dir,
  catalog_type = "self-contained",
  overwrite = TRUE
)
```

The resulting directory structure looks like:

```
catalog/
  catalog.json
  terrain/
    collection.json
    dem-001/
      dem-001.json
```

We can read it back:

```{r}
cat_read <- read_stac(file.path(catalog_dir, "catalog.json"))
coll_read <- read_stac(file.path(catalog_dir, "terrain", "collection.json"))
item_read <- read_stac(file.path(
  catalog_dir,
  "terrain",
  "dem-001",
  "dem-001.json"
))

item_read
```

## Serving the catalog as a STAC API

A static catalog on disk is enough for many uses, but a searchable STAC API
needs a database behind it. That is the job of the companion package
[stacserver](https://github.com/stevenpawley/stacserver), which ingests the
collections and items built here into PostgreSQL/PostGIS and serves them
through a `plumber` router. See its "Serving a STAC API" vignette.
